---
title: "ANOVA with R"
author: "Valentin Stefan"
date: "Summer Schools on Pollination Ecology, 2018"
output:
  ioslides_presentation:
    smaller: yes
    widescreen: yes
---

## Goals for this lab

- Understand why we use ANOVA to answer scientific questions
- Learn how to use ANOVA in R


## Motivation, PD & PL - 1

In the field, you conducted **breeding system** and **pollen supplementation experiments**. 

**Breeding system experiments: measure pollinator dependence (PD)**

- quantify the reliance of a plant species on pollinators for reproduction

**Supplementation experiments: measure pollen limitation (PL)**

- quantify the magnitude by which plant reproduction is limited by pollen receipt from pollinators

Use ANOVA to ask whether plants are **pollinator dependent** and/or **pollen limited** (compare *bagged*, *control* and *supplement* treatments).


## Motivation, PD & PL - 2

Situations you may encounter when measuring reproductive success:

```{r, echo=FALSE, out.width="85%", fig.align='center'}
library(data.table)

generate_data <- function(n, mean_bag, mean_con, mean_sup, sd, title = ""){
  set.seed(321)
  dt <- data.table(bag = rnorm(n = n, mean = mean_bag, sd = sd),
                   con = rnorm(n = n, mean = mean_con, sd = sd),
                   sup = rnorm(n = n, mean = mean_sup, sd = sd))
  dt2 <- melt(dt, measure.vars = 1:3)
  dt2[, value := abs(value)]
  boxplot(value ~ variable, 
          data = dt2, 
          outline = FALSE, 
          main = title, 
          cex.axis = 1.5)
  stripchart(value ~ variable, data = dt2,
             method = "jitter",
             jitter = 0.1,
             pch = 21,
             col = rgb(1,0,0, alpha = 0.6),
             cex = 1.5,
             vertical = TRUE,
             add = TRUE)
}
par(mfrow = c(1,3), mar = c(b=2,l=3,t=1,r=0))
generate_data(n = 50, mean_bag = 0.1, mean_con = 0.5, mean_sup = 0.5, sd = 0.15)
generate_data(n = 50, mean_bag = 0.5, mean_con = 0.5, mean_sup = 0.5, sd = 0.15)
generate_data(n = 50, mean_bag = 0.1, mean_con = 0.35, mean_sup = 0.5, sd = 0.15)
```

## Motivation, PD & PL - 3

**Pollinator dependence (PD)** & **pollen limitation (PL)** based on a reproductive success measure (e.g ratio between viable seeds and total number of ovules)

|pollinator dependence (PD)|pollen limitation (PL)    |
|--------------------------|--------------------------|
|if bag < sup              |if con < sup              |

`

**Auto-fertility index (AFI)**

- AFI is the ratio of average reproductive success in bagged vs. supplement. 
- PD if AFI < 0.2

**Magnitude (effect size) of PL (the log response ratio)**

- `ES = ln(reproductive success in supplement treatment) - ln(reproductive success in control treatment)`
- A low value (close to 0) indicates no PL


## Motivation, PD & PL - 3

|pollinator dependence (PD)|pollen limitation (PL)|
|--------------------------|----------------------|
|if bag < sup              |if con < sup          |

```{r, echo=FALSE, out.width="75%", fig.align='center'}
par(mfrow = c(1,3), mar = c(b=2,l=3,t=1,r=0))
generate_data(n = 50, mean_bag = 0.1, mean_con = 0.5, mean_sup = 0.5, sd = 0.15, 
              title = "PD & not PL")
generate_data(n = 50, mean_bag = 0.5, mean_con = 0.5, mean_sup = 0.5, sd = 0.15,
              title = "not PD, nor PL")
generate_data(n = 50, mean_bag = 0.1, mean_con = 0.35, mean_sup = 0.5, sd = 0.15,
              title = "PD & PL")
```

## ANalysis Of VAriance (ANOVA)

How many means do you compare?

- 2 means - **t-test**

- more than 2 means - **ANOVA** (bag, cont, sup)

Note - can also use ANOVA for two means. But it is standard practice to use t-test in such cases.

ANOVA 

- **Null Hypothesis (H0)**: no difference between the group means

- **Alternative hypothesis (H1)**: means are different (do further tests to see which ones)

Next - an example with R.

## Outline of the exercise

- Read data
- Check assumptions (because ANOVA is a "parametric" test)
  - normality
  - homogeneity of variance
- ANOVA summary: are there differences or not?
- "post hoc" tests: where are the differences?
